From Reversible Programs to \\ Univalent Universes and Back

### 2017-05-26 upi.tex
expand intro to Pi subsection.  Will next work on not introducing a new syntax...

\usepackage{bbold}
\usepackage{bussproofs}
\usepackage{keystroke}
\usepackage{comment}
\usepackage{tikz}

@@

The practice of programming languages is replete with \emph{ad hoc} instances of
reversible computations such as database transactions, mechanisms for data
provenance, checkpoints, stack and exception traces, logs, backups, rollback
recoveries, version control systems, reverse engineering, software transactional
memories, continuations, backtracking search, and multiple-level ``undo''
features in commercial applications. In the early nineties,
Baker~\cite{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
treatment of reversibility but intensive research in full-fledged reversible
models of computations and reversible programming languages was only sparked by
the discovery of deep connections between physics and
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275}
and by the potential for efficient quantum
computation~\cite{springerlink:10.1007/BF02650179}.

The early developments of reversible programming languages started with a
conventional programming language, e.g., an extended $\lambda$-calculus, and either
(i) extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible} or
(ii) imposed constraints on the control flow constructs to make them
reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}. More modern approaches
recognize that reversible programming languages require a fresh approach and
should be designed from first principles without the detour via conventional
irreversible languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}.

@@

In previous work, Carette, Bowman, James, and
Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
the~$\Pi$ family of reversible languages based on type isomorphisms and
commutative semiring identities. The fragment without recursive types is
universal for reversible boolean circuits~\cite{James:2012:IE:2103656.2103667}
and the extension with recursive types and trace
operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
language~\cite{James:2012:IE:2103656.2103667,rc2011}. In this paper we will
identify a ``featherweight'' representative to use for the formal
development. But first, we illustrate the flavor of the family of languages with
some examples written in a Haskell syntax. First consider the following program:

>>>>

\usepackage{bbold}
\usepackage{bussproofs}
\usepackage{keystroke}
\usepackage{comment}
\usepackage{tikz}
\usepackage[inline]{enumitem}

@@

The practice of programming languages is replete with \emph{ad hoc} instances of
reversible computations: database transactions, mechanisms for data
provenance, checkpoints, stack and exception traces, logs, backups, rollback
recoveries, version control systems, reverse engineering, software transactional
memories, continuations, backtracking search, and multiple-level ``undo''
features in commercial applications. In the early nineties,
Baker~\cite{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
treatment of reversibility. But intensive research in full-fledged reversible
models of computations and reversible programming languages was only sparked by
the discovery of deep connections between physics and
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
and by the potential for efficient quantum
computation~\cite{springerlink:10.1007/BF02650179}.

The early developments of reversible programming languages started with a
conventional programming language, e.g., an extended $\lambda$-calculus, and either
\begin{enumerate*}[(i)]
\item extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or
\item imposed constraints on the control flow constructs to make them
reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}. 
\end{enumerate*}
More modern approaches recognize that reversible programming languages require
a fresh approach and should be designed from first principles without the
detour via conventional irreversible
languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}.

@@

In previous work, Carette, Bowman, James, and
Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
the~$\Pi$ family of reversible languages based on type isomorphisms and
commutative semiring identities. The fragment without recursive types is
universal for reversible boolean circuits~\cite{James:2012:IE:2103656.2103667}
and the extension with recursive types and trace
operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
language~\cite{James:2012:IE:2103656.2103667,rc2011}. While at first sight,
$\Pi$ might appear \emph{ad hoc},~\cite{Carette2016} shows that it arises naturally
from an ``extended'' view of the Curry-Howard Correspondance: rather than looking
at mere \emph{inhabitation} as the main source of analogy between logic
and computation, \emph{type equivalences} becomes the source of analogy.  This
allows one to see an analogy between algebra and reversible computation.
Furthermore, this works at multiple levels: that of $1$-algebra (types form a
semiring under isomorphism), but also $2$-algebra (types and equivalences form a 
weak Rig Groupoid).  In other words, by taking ``weak Rig Groupoid'' as the 
starting semantics, one naturally gets $\Pi$ as the syntax for the language of
proofs of isomorphisms -- in the same way that many terms of the
$\lambda$-calculus arise from Cartesian Closed Categories.

By restricting to finite types, two operational semantics emerge:
\begin{enumerate*}{(i)}
\item a (reversible) operational semantics of programs over finite types, and
\item a (reversible) operational semantics of program transformers of such programs.
\end{enumerate*}
Such programs can be interpreted as generalized permutations, as they work on
structure as well as the underlying skeleton of finite types, aka the natural numbers.
Many of the transformers can be \emph{oriented}, and thus be interpreted as
\emph{optimizations} in some sense.

In this paper we will
identify a ``featherweight'' representative to use for the formal
development. But first, we illustrate the flavor of the family of languages with
some examples written in a Haskell syntax. First consider the following program:
